---
title: "Edbuild Minnesota Analysis"
author: "Krista Kaput"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}

# load ---------------

# this will prevent r code chunks from printing to final doc
knitr::opts_chunk$set(echo = FALSE)


options(scipen = 999)

library(tidyverse)
library(sf)
library(viridis)
library(scales)

source("edbuild_clean.R")
source("edbuild_map_data.R")


# clean -----------

mn_dist_map <- mn_sd_map |> 
  merge(mn_dist, by.x = "geoid", by.y = "ncesid")

```

## Data Sources

The EdBuild data for Minnesota was pulled from the `edbuildr` package on 2023-05-08.

Map data for Minnesota districts was pulled from the `edbuilmapr` package on 2023-05-08.

More detail on both packages can be found on [EdBuild's website](http://viz.edbuild.org/workshops/edbuildr/).

## Overview of Minnesota district data

Using the `summary()` function, we can see a quick preview of the district data from the `edbuildr` package.

```{r mn-summary}
summary(mn_dist_map)
```

## Mapping Minnesota data


### Poverty 

The distrcts with the highest poverty rates in Minnesota are in the Twin Cities and on reservations. 

```{r mn-poverty}

ggplot(mn_dist_map, aes(fill = stpovrate)) +
  geom_sf() + 
  scale_fill_viridis(labels = label_percent()) +
  labs(fill = "Census poverty rate") +
  theme_void()

```

### Urbanicity

The NCES urbanicity definitions are *pretty* granular.

```{r mn-urban}

ggplot(mn_dist_map, aes(fill = durbanicity)) +
  geom_sf() + 
  scale_fill_viridis_d() +
  labs(fill = "NCES Urbanicity") +
  theme_void()

```

If we **simplifiy** to four urbanicity categories, the high frequency of rural districts becomes more apparent.

```{r mn-simple-urban}

ggplot(mn_dist_map, aes(fill = urbanicity)) +
  geom_sf() + 
  scale_fill_viridis_d() +
  labs(fill = "Simplified Urbanicity") +
  theme_void()

```


## Local Revenue

There aren't any clear outliers. However, local per-pupil revenues greatly varies from 701 in Red Lake Public Schools to 10,416 in Ivanhoe Public School District. 
```{r local-rev}
ggplot(mn_dist_map, aes(fill = lrpp)) +
  geom_sf()+
  scale_fill_viridis(labels = label_dollar()) +
  labs(fill = "Local Revenue Per-Pupil")+
  theme_void()

```


## Urbanicity and local revenue

Rural districts generate **significantly** less revenue than urban or suburban districts in Minnesota. 

```{r local-rev-urban}

urban_rev_summary <- mn_dist |> 
  # get average local revenue per pupil by urbanicity
  group_by(urbanicity) |> 
  summarise(lrpp_avg = mean(lrpp, na.rm = T)) |> 
  # format summary as currency, rounded to the nearest dollar
  mutate(lrpp_avg = dollar(lrpp_avg, accuracy = 1)) |> 
  rename(Urbanicity = urbanicity,
         `Average Local Rev. Per-Pupil` = lrpp_avg)

knitr::kable(urban_rev_summary)

```