---
title: "Basic Data Analysis and Visualization in R"
date: "June 13, 2023"
format: 
  revealjs:
    chalkboard: true
execute:
  freeze: auto
  echo: false
---


# Homework Review

## Solving a common problem with the homework


# Introduction to the tidyverse

## R is a functional programming language - most of what you will do in R is work with functions 

A function: 

-  Takes a specified input
-  Performs an operation 
-  Returns an output

Functions are helpful tools to reduce repetition in typing and improves your code's 
consistency, reliability and readability. 

## The tidyverse is a powerful collection of R packages - collections of functions - that work well together (1/2)

- The most popular packages in the R community are past of what is called the "tidyverse," which includes packages like ggplot2, tidyr, stringr, tibble, purrr
- Tidyverse packages are built to work together. Every tidyverse package contains functons that can manipulate or visualize data that lives in data frames. 

## The tidyverse is a powerful collection of R packages - collections of functions - that work well together (2/2)

- Most functions in the tidyverse require a data frame (R's version of a table) as the first argument in each function. The functions can be "chained" together with other functions. 

We will explore the tidyverse by looking at EdBuild's FY2019 education data compiled from the F33 survey, SAIPE, EDGE, and CCD. The edbuildr package provides access to clean district
data on funding, student demographics, and wealth.

## To read EdBuild's finance data into our working environment we will use the masterpull() function 

masterpull(data_year = 2019, data_type = “fin”): Downloads a clean dataset from EdBuild’s server. There are three options for the data_type argument:

- “geo”: only includes districts with geographic boundaries
(no charter/special school districts) that meet EdBuild’s
criteria for fiscal analysis
- “fin”: includes all districts that meet EdBuild’s criteria
for fiscal analysis
- “full”: includes all districts – TO BE USED WITH
CAUTION


## The dplyr() package provides many functions to manipulate your data frames' columns and rows 

The functions you'll most frequently use from the dplyr packages are: 

- select(): names columns to keep from a data frame
- rename(): name columns to keep from a data frame 
- filter(): remove rows that do not meet the condition in the logical statement from the output
- mutate(): create a new column that will be added to the end of your data frame. 

# Iterative exploratory analysis 

## LIVE CODING EXAMPLE: exploring_the_tidyverse.R 

```{r}
#| echo: true

# load ----------

options(scipen = 999)

library(tidyverse)
library(edbuildr)

# Krista likes to denote which spreadsheet is raw for transparency 
dist_fy19_raw <- masterpull(data_type = "geo")

# filter Minnesota data and clean -----

# filter, rename, and mutate data for Minnesota 
mn_ed_data <- dist_fy19_raw |>
  filter(State == "Minnesota") |>
  rename(district = NAME,
         county = County,
         enroll = ENROLL, 
         total_local_rev = LR,
         total_state_rev = SR,
         total_state_local_rev = SLR, 
         urbanicity = dUrbanicity,
         operational_schools = dOperational_schools, 
         district_type = dType, 
         white_enroll = dWhite, 
         sped_enroll = dIEP, 
         ell_enroll = dLEP, 
         econ_dis_enroll = StPov,
         bipoc_pct = pctNonwhite, 
         pov_pct = StPovRate,
         median_house_income = MHI, 
         median_prop_value = MPV) |>
  mutate(bipoc_enroll = enroll - white_enroll,
         ell_pct = ell_enroll/enroll, 
         sped_pct = sped_enroll/enroll, 
         loc_rev_pp = total_local_rev/enroll,
         state_rev_pp = total_state_rev/enroll,
         local_state_rev_pp = total_state_local_rev/enroll) |>
  select(district, county, enroll, loc_rev_pp, state_rev_pp, local_state_rev_pp, 
         total_local_rev, total_state_rev, total_state_local_rev, urbanicity, 
         operational_schools, district_type, pov_pct, bipoc_pct, ell_pct, sped_pct)

```

# Break 

# Visual exploratory analysis 

## Summarizing columns can be helpful, but it can hide nuance in your data that can be better seen via plotting 

The ggplot2 package is the most widely-used data visualization approach in the R ecosystem. 

- ggplot plots are created by starting with your data, then building up layers 
- Once your specify your data, you can add layers of "geoms" to create your plot 
- Today we will use geom_histogram() and geom_point()

# Iterating your way to beauty with ggplot2 using Minnesota data

## Step 1: Create a basic plot 

```{r}
#| echo: true
library(tidyverse)

# first minnesota plot 
ggplot(mn_ed_data, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point()

```

## Step 2: Address missing values

We need to figure out what’s going on with our missing values before we go any further!


```{r}
#| echo: true
# investigate missing values 
mn_na_dist <- mn_ed_data |>
  filter(is.na(state_rev_pp) | is.na(pov_pct)) |>
  arrange(district_type) 

view(mn_na_dist)

```

The missing data makes sense. All eight of the rows don't have any available data. 

## Step 3: Clean up formatting of chart elements (1/5)

Now that we understand our missing data, we can create an updated dataframe and re-plot our newly cleaned data.

```{r}
#| echo: true
# create tidy df
mn_ed_clean <- mn_ed_data |>
  filter(!is.na(state_rev_pp), 
         !is.na(pov_pct))

view(mn_ed_clean)

# Drop the 2 districts that have no enrollment
mn_ed_clean <- mn_ed_clean |>
  filter(enroll > 0)

# first minnesota plot w/ clean data
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point()


```

## Step 3: Clean up formatting of chart elements (2/5)

We see some overlap in the points. Reducing the opacity of the points can be accomplished by setting the alpha parameter in geom_point() to a value less than 1. Setting it to .5 will make data points 50% translucent.

```{r}
#| echo: true
# reduce opacity of points
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point(alpha = .5)


```

## Step 3: Clean up formatting of chart elements (3/5)
Let’s take care of some formatting issues. Our axes don’t look great - the decimals ought to be percentages. Here, the scales package provides some help.

```{r}
#| echo: true
# format axes 
library(scales)

# format the x and y axes
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point(alpha = .5) + 
# make sure you have the `scales` package loaded!
  scale_x_continuous(labels = percent_format(accuracy = 1)) 


```

## Step 3: Clean up formatting of chart elements (4/5)
Next, we should add some labels to our axes that make sense, along with a title for our plot and a caption that details our data sources.

```{r}
#| echo: true

# add data labels 
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point(alpha = .5) + 
# make sure you have the `scales` package loaded!
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019") 

```
## Step 3: Clean up formatting of chart elements (5/5)
Themes can be used to change the appearance of elements in your plot. There are many stock options, but I prefer theme_bw() for its clean appearance and helpful and unobtrusive grid lines.

```{r}
#| echo: true

# change theme
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019") +
  theme_bw()


```

# Break 

## Step 4: Add a new layer of data 
Now that we have a decent-looking graph, let’s add in a new data element to vary point size by enrollment.

```{r}
#| echo: true

# add size element 
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019") +
  theme_bw()


```

## Step 5: Tidy up formatting (1/2)
Adding a new variable for size creates a legend. We need to tidy the legend’s labels and the title.
```{r}
#| echo: true

# clean up the legend
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  # change legend label formatting
  scale_size(labels = comma) +
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```
## Step 5: Tidy up formatting (2/2)
We can also adjust some paramenters to allow for more visual contrast in size. By default, ggplot2 will adjust points’ radii based on the size variable. Using area is a more visually honest way to represent the data, so let’s make that change.

```{r}
#| echo: true

# create more contrast in size
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  # change size scaling to vary by area, not radius + change max size
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()

```

## Step 6: Repeat steps 4-5 as needed (1/5)
Adding color can be helpful. Let's add color based on urbanicity. 
```{r}
#| echo: true

# add in color based on urbanicity
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()

```
## Step 6: Repeat steps 4-5 as needed (2/5)
That doesn't look great. Let’s clean up the data a little more to reduce 11 urbanicity levels to four, then tidy up the legend label.
```{r}
#| echo: true

# clean up the unicorn vomit
mn_ed_clean <- mn_ed_clean |>
  mutate(urbanicity = fct_collapse(as.factor(urbanicity),
                                   City = c("11-City: Large", 
                                            "12-City: Mid-size",
                                            "13-City: Small"),
                                   Suburb = c("21-Suburb: Large",
                                              "22-Suburb: Mid-size",
                                             "23-Suburb: Small"),
                                   Town = c("31-Town: Fringe",
                                            "32-Town: Distant",
                                            "33-Town: Remote"),
                                   Rural = c("41-Rural: Fringe",
                                             "42-Rural: Distant",
                                             "43-Rural: Remote")))

# better colors and tidy legend label
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```

## Step 6: Repeat steps 4-5 as needed (3/5)
We can and should adjust the colors used. R recognizes some pretty funky color names, which can be found in this helpful cheat sheet.
```{r}
#| echo: true
# adjust colors manually
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
    # create manual color palette
  # color names pulled from a pdf y'all should bookmark
  # http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
  scale_color_manual(values = c("tomato3", "steelblue2",
                                "seagreen3", "orchid1")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```


## Step 6: Repeat steps 4-5 as needed (4/5)
We should strive to make our analyses as accessible as possible. The viridis package includes some color palettes that are friendly for folks with color blindness, which affects 5-10 percent of the US population.
```{r}
#| echo: true
# use colors better for visual impairments
library(viridis)
# adjust colors manually

ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
  # use a colorblind-friendly palette
  scale_color_viridis_d() +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```
## Step 6: Repeat steps 4-5 as needed (5/5)
Let’s adjust the range of colors used to exclude that hard-to-see yellow.
```{r}
#| echo: true
# that yellow is hard to see - let's adjust the range
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
  # adjust color range
  scale_color_viridis_d(end = .8) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```
## Step 7: Replace a layer of data
Sometimes color isn’t the answer - let’s try facets instead.
```{r}
#| echo: true
# use facets instead of color
ggplot(mn_ed_clean, aes(x = pov_pct, y = state_rev_pp, size = enroll,
                        color = urbanicity)) +
  geom_point(alpha = .5) + 
  # add facet by urbanicity
  facet_wrap(~urbanicity) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_area(labels = comma, max_size = 10) + 
    labs(x = "Student Poverty Rate", y = "State Per-Pupil Revenue",
       title = "State Per-Pupil Revenue by Student Povert Rate in Minnesota School Districts",
       caption = "Source: Edbuild Data, 2019",
       # add nice label for size element
  size = "Enrollment") +
  theme_bw()


```
# Using histograms to compare state and local revenue 

## Use Edbuildr data to create state summaries 
```{r}

# summarise ---------

# create a state summary dataframe
state_summary <- dist_fy19_raw |> 
  group_by(State) |> 
  summarise(enroll_med = median(ENROLL, na.rm = T),
            n_schools_med = median(dOperational_schools, na.rm = T),
            nonwhite_pct_med = median(pctNonwhite, na.rm = T),
            st_pov_rate_med = median(StPovRate, na.rm = T),
            sd_area_med = median(sd_area, na.rm = T),
            student_per_sq_mile_med = median(student_per_sq_mile, na.rm = T),
            mhi_med = median(MHI, na.rm = T),
            mpv_med = median(MPV, na.rm = T),
            state_loc_rev_pp_med = median(SLRPP, na.rm = T))

```

## Create histograms 
```{r}
# histogram of median property value
ggplot(data = state_summary) +
  geom_histogram(mapping = aes(x = mpv_med), binwidth = 40000)

# histogram of median household income
ggplot(data = state_summary) +
  geom_histogram(mapping = aes(x = mhi_med), binwidth = 3000)

# histogram of state + local pp rev
ggplot(data = state_summary) +
  geom_histogram(mapping = aes(x = state_loc_rev_pp_med), binwidth = 2000)

```
# Homework assignment 





